---
title: "Lightning simulation"
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# Task

- Create a random lightning algorithm, and visualize the result!
- The lightning should be a random walk from top to bottom.
- At each step, there is a 1% chance of forking.
- Each fork in itself is a new random walk.
- The forks do not need to fork further.
- Visualize the lightning, preferably using black background and blue/white foreground color. 
- Try to do this using no loops!

# Dataset

There is no dataset for this project. You have to simulate from scratch!

```{r}
library(tidyverse)

lightning_length = 500 # How tall the lightning is (in points)
forking = .01 # Chance of forking at each step
p_forking = c(1 - forking, forking) # The probability of not forking and forking

lightning <-
    # Creating the data structure
    tibble(y = 1:lightning_length,
           # The position of the lightning on the x scale changes randomly
           change = sample(-1:1, size = lightning_length, replace = TRUE),
           # Decide if there is a forking at that point
           fork = sample(0:1, prob = p_forking, size = lightning_length, replace = TRUE),
           # If forking, name the new fork with adding a number to the name
           fork_name = if_else(fork > 0, paste0("fork_", cumsum(fork)), NA_character_)) %>% 
    pivot_wider(names_from = "fork_name", values_from = "fork") %>%
    # A variable called NA was also created we don't need it
    select(-`NA`) %>% 
    # Fill all fork variables from the point where it is not NA
    fill(matches("fork_\\d+")) %>% 
    # From now on, calculate the actual x position of each for
    # Up until now, we only calculated the relative changes
    mutate(fork_zero = cumsum(change),
           across(matches("fork_\\d+"), 
                  ~if_else(.x == 1, sample(-1:1, size = lightning_length, replace = TRUE), 0L)),
           across(matches("fork_\\d+"), 
                  ~if_else(is.na(.x), 0L, .x) %>% 
                      cumsum() + fork_zero)) %>% 
    # Cast into long format for plotting
    pivot_longer(matches("fork_\\d+"), 
                 names_to = "fork_no",
                 values_to = "x")

# Visualize the results
lightning %>% 
    ggplot() +
    aes(y = rev(y), x = x, group = fork_no) +
    # We create 3 lightnings wit decreasing width and increasing light (not necessary)
    geom_ribbon(aes(xmin = x - 1, xmax = x + 1), color = "darkblue") +
    geom_ribbon(aes(xmin = x - .5, xmax = x + .5), color = "lightblue", alpha = .5) +
    geom_ribbon(aes(xmin = x - .1, xmax = x + .1), color = "white", alpha = .5) +
    coord_cartesian(xlim = c(-150, 150)) +
    # Remove axes and add black background
    theme_void() +
    theme(plot.background = element_rect(fill = "black")) 
```

